---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
First, we'll load the libraries and dependencies which will be required for processing and visualization.

```{r install dependencies, echo=TRUE}
list.of.packages <- c("tidyverse","ggplot2","lubridate","forcats")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[, "Package"])]
if(length(new.packages)) install.packages(new.packages)
library(tidyverse); library(ggplot2); library(lubridate); library(forcats)
```

Then, we'll read in the csv data as a dataframe using the tidyverse::readr package. 

```{r read data, echo=TRUE}
unzip("activity.zip")
activity <- read_csv("activity.csv")
```

## What is mean total number of steps taken per day?

Calculate the total number of steps taken each day
```{r steps by day, echo=TRUE}
stepsByDay <- activity %>% group_by(date) %>% summarise(steps=sum(steps))
```

Create a histogram of Total Number of Steps Taken Per Day
```{r histogram of steps per day, echo=TRUE, fig.height=8, fig.width=8}
ggplot(data=stepsByDay) + geom_histogram(mapping=aes(x=steps), fill="red", color="white", bins=30, na.rm=TRUE) + xlab("Steps") + ylab("Count") + ggtitle("Total Number of Steps Taken Per Day")
```

Calculate and report the mean of total number of steps taken per day
```{r mean steps, echo=TRUE}
meanStepsByDay <- stepsByDay %>% summarise(steps=mean(steps, na.rm=TRUE))
meanStepsByDay 
```

Calculate and report the median number of steps taken per day
```{r median steps, echo=TRUE}
medianStepsByDay <- stepsByDay %>% summarise(MedianDailySteps=median(steps, na.rm=TRUE))
medianStepsByDay
```

## What is the average daily activity pattern?

Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken averaged across all days (y-axis)
```{r interval average, echo=TRUE, fig.height=8, fig.width=8}
intActivity <- activity %>% group_by(interval) %>% summarise(stepsAverage=mean(steps, na.rm=TRUE))
ggplot(intActivity, aes(interval, stepsAverage)) + geom_line() + xlab("Time") + ylab("Average Number of Steps") + scale_x_continuous(breaks=c(0,500,1000,1500,2000), labels=c("0:00","5:00","10:00","15:00","20:00"))
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r interval max, echo=TRUE}
intActivity %>% filter(stepsAverage==max(stepsAverage)) 
```

## Imputing missing values

Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r rows with missing values, echo=TRUE}
activity %>% summarise(steps=sum(is.na(steps)))
```

Use the mean number of steps for an interval to fill all of the missing values in the dataset.
Create a new dataset that is equal to the original dataset, but with the missing data filled in.
```{r impute missing data with mean(), echo=TRUE}
imputeSteps <- activity %>% group_by(interval) %>% mutate(steps=ifelse(is.na(steps),mean(steps, na.rm=TRUE),steps))
```

```{r created new imputed dataset, echo=TRUE}
imputedStepsByDay <- imputeSteps %>% group_by(date) %>% summarise(steps=sum(steps))
```

Make a histogram of the total number of steps taken each day.
```{r imputed steps by data histogram, echo=TRUE, fig.height=8, fig.width=8}
ggplot(data=imputedStepsByDay) + geom_histogram(mapping=aes(x=steps), fill="red", color="white", bins=30, na.rm=TRUE) + xlab("Steps") + ylab("Count") + ggtitle("Total Number of Steps Taken Per Day (with Interval Mean Imputed for NAs)")
```

Calculate and report the mean number of steps taken per day.
```{r imputed mean steps per day, echo=TRUE}
impMeanStepsByDay <- imputedStepsByDay %>% summarise(MeanDailySteps=mean(steps))
impMeanStepsByDay
```

Calculate and report the median number of steps taken per day.

```{r imputed median steps per day, echo=TRUE}
impMedianStepsByDay <- imputedStepsByDay %>% summarise(MedianDailySteps=median(steps))
impMedianStepsByDay
```

Calculate the difference between imputed mean and mean values where NAs are dropped.

```{r difference imputed mean and mean}
impMeanStepsByDay - meanStepsByDay
```

Note that there is no difference between the imputed mean values and mean values where NAs are dropped.

Calculate the difference between imputed mean and mean values where NAs are dropped. 
```{r difference imputed median and median}
impMedianStepsByDay - medianStepsByDay
```

There is a 1+/- step difference when using imputed mean values for NA values. This is practically neglible.

## Are there differences in activity patterns between weekdays and weekends?
First, convert date to an ordered factor variable of days with two levels: weekend and weekday.
```{r, create weekday and weekend factor, echo=TRUE}
activityDays <- imputeSteps %>% mutate(day=wday(date, label=TRUE, abbr=FALSE)) 
activityDaysFactor <- activityDays %>% mutate(daytype=fct_collapse(day, weekend=c("Sunday", "Saturday"), weekday=c("Monday","Tuesday","Wednesday", "Thursday","Friday")))
```

Then make a panel plot containing a time-series plot of average steps taken averaged across all weekdays and weekends.
```{r, create weekday and weekend panel plot, echo=TRUE, fig.height=8, fig.width=8}
intervalWeekendWeekday <- activityDaysFactor %>% group_by(interval, daytype) %>% summarise(steps=mean(steps))
ggplot(intervalWeekendWeekday, aes(interval, steps)) + geom_line() + facet_wrap(~ daytype, nrow=2) + xlab("Time") + ylab("Average Number of Steps") + scale_x_continuous(breaks=c(0,500,1000,1500,2000), labels=c("0:00","5:00","10:00","15:00","20:00"))
```
